---
title: "Pacing Strategy Analysis Plan"
output: html_document
---

# Goal

The broad goal is to understand how pacing strategy relates to final race outcome.

We have sector-level race data for each runner, plus pre-race information such as
`runner_overall_index`, which can be treated as a rough estimate of running ability
before the race. The main research question is:

> Among runners with similar pre-race ability, which pacing patterns are associated
> with better or worse outcomes?

This is important because absolute race time alone mixes many things together:
fitness, weather, course difficulty, injury, nutrition, and tactical pacing. A strong
runner can run a fast time even with bad pacing, while a weaker runner can execute a
very good race and still finish later in absolute terms. The analysis should therefore
compare runners mostly against other runners of similar expected ability.

# Main Intuition

The raw question "who was fastest?" is not enough. A better question is:

> Given what we expected from this runner before the race, did they perform better
> or worse than expected, and is that related to how they distributed effort across
> the race?

For example, two runners can have the same final time but very different pacing:

- Runner A starts fast, slows down heavily, and survives to the finish.
- Runner B starts conservatively, keeps effort steady, and finishes strongly.

Even if both finish in the same time, their pacing strategies tell different stories.
If Runner B had a worse pre-race index, Runner B may actually have executed the better
race.

# Important Columns

Likely key columns:

- `runner_id`: unique runner/race/year identifier.
- `race_id`: race identifier.
- `year`: race year.
- `runner_overall_index`: pre-race ability estimate.
- `runner_final_time`: final race time, if available for finishers.
- `runner_final_status` or similar: finisher, DNF, DNS, etc.
- `sector_checkpoint_id`: ordered checkpoint or sector index.
- `sector_time`: time spent in each sector.
- `sector_time_min`: sector time converted to minutes.
- `sector_race_time_at_checkpoint`: cumulative race time at each checkpoint.
- `sector_race_time_at_checkpoint_min`: cumulative race time in minutes.

Before modeling, we should confirm the exact schema and units. In particular:

- Is `runner_overall_index` lower for stronger runners, or higher for stronger runners?
- Are all sectors available for all finishers?
- How are DNF runners represented after their final reached checkpoint?
- Are checkpoint IDs now normalized to clean `0, 1, 2, ..., K` values within each
  `race_id` and `year`?

# Unit Of Analysis

There are two natural analysis levels.

## Runner-Level Analysis

One row per runner. This is useful for modeling final outcome.

Example features:

- final race time
- DNF indicator
- pre-race overall index
- pacing summary features
- slowdown features
- early/middle/late split ratios

This level answers:

> Which pacing summaries are associated with better final outcomes?

## Runner-Sector-Level Analysis

One row per runner per sector. This is useful for understanding how the race unfolds.

Example features:

- sector time
- sector rank
- cumulative time
- sector number
- relative sector pace
- position change from sector to sector

This level answers:

> Where in the race do different groups gain or lose time?

# Ability Groups

Because `runner_overall_index` is a pre-race ability estimate, we can use it to group
runners into comparable ability bands.

Possible groupings:

- Quantiles, for example 5 groups or 10 groups.
- Fixed bins if the index has an interpretable scale.
- Gender-specific or category-specific quantiles if the distribution differs strongly
  by group.

A simple starting point:

```{r}
# Example idea in pandas syntax, not R:
# runner_df["ability_group"] = pd.qcut(
#     runner_df["runner_overall_index"],
#     q=5,
#     labels=["G1", "G2", "G3", "G4", "G5"]
# )
```

If lower `runner_overall_index` means stronger runner, then group `G1` would contain
the strongest runners. If higher means stronger, then the labels should be reversed.

# Normalizing Final Outcome

Absolute final time is not enough because stronger runners are expected to finish
faster. We need an outcome that says whether a runner performed better or worse than
expected.

## Option 1: Compare Within Ability Group

For each ability group, compute each runner's performance relative to the group.

Let:

$$T_i = \text{final time of runner } i$$

and:

$$\bar{T}_{g(i)} = \text{average final time among finishers in runner } i\text{'s ability group}$$

Then:

$$\Delta T_i = T_i - \bar{T}_{g(i)}$$

Interpretation:

- $\Delta T_i < 0$: runner was faster than the average runner in their ability group.
- $\Delta T_i > 0$: runner was slower than the average runner in their ability group.

This is easy to explain, but it can be rough if ability groups are too wide.

## Option 2: Regression Residual

Fit a baseline model that predicts final time from pre-race ability.

For example:

$$T_i = \beta_0 + \beta_1 \cdot I_i + \epsilon_i$$

where:

- $T_i$ is final race time.
- $I_i$ is `runner_overall_index`.
- $\epsilon_i$ is the residual.

Then:

$$\epsilon_i = T_i - \hat{T}_i$$

Interpretation:

- $\epsilon_i < 0$: faster than expected.
- $\epsilon_i > 0$: slower than expected.

This is attractive because it creates a clean "performance relative to expectation"
metric. We can later improve the baseline model with gender, age category, year, or
race-specific effects.

## Option 3: Percentile Outcome Within Ability Group

Within each ability group, rank finishers by final race time.

For runner $i$:

$$P_i = \text{percentile rank of final time within ability group}$$

Interpretation:

- Low percentile can mean a strong result.
- High percentile can mean a weak result.

This is robust and easy to communicate, but it loses some information about how large
the time gaps are.

# Normalizing Pacing

Pacing should also be normalized. Absolute sector times are not comparable across
runners because stronger runners are faster everywhere.

The useful question is not:

> Was this sector fast in absolute minutes?

The useful question is:

> Was this sector fast relative to this runner's own race, or relative to similar
> runners?

## Runner-Relative Sector Share

For each runner and sector:

$$s_{ij} = \frac{t_{ij}}{T_i}$$

where:

- $t_{ij}$ is runner $i$'s time in sector $j$.
- $T_i$ is runner $i$'s final race time.

Interpretation:

- $s_{ij}$ is the fraction of total race time spent in sector $j$.
- This creates a pacing profile independent of total speed.

This is good for finishers. For DNF runners, total final time is not comparable, so a
different treatment is needed.

## Group-Relative Sector Performance

Within each `race_id`, `year`, `ability_group`, and `sector_checkpoint_id`, compare a
runner's sector time to the median sector time for that group.

Let:

$$m_{g,j} = \text{median sector time for ability group } g \text{ in sector } j$$

Then:

$$r_{ij} = \frac{t_{ij}}{m_{g(i),j}}$$

Interpretation:

- $r_{ij} < 1$: faster than the median comparable runner in that sector.
- $r_{ij} > 1$: slower than the median comparable runner in that sector.

This is useful because it adjusts for sector difficulty. A climb sector and a descent
sector should not be compared on the same absolute scale.

## Cumulative Relative Time

We can also compare cumulative time at each checkpoint:

$$c_{ij} = \frac{C_{ij}}{\hat{C}_{g(i),j}}$$

where:

- $C_{ij}$ is runner $i$'s cumulative time at checkpoint $j$.
- $\hat{C}_{g(i),j}$ is the median cumulative time for that ability group at checkpoint
  $j$.

This helps show whether a runner is ahead of or behind the expected trajectory at each
point in the race.

# Pacing Strategy Features

Once sector times are normalized, we can create interpretable pacing features.

## Early, Middle, Late Split Ratios

Divide the race into thirds:

- early race
- middle race
- late race

Then compute average relative pace in each part.

Example:

$$E_i = \text{mean relative sector time in early sectors}$$

$$M_i = \text{mean relative sector time in middle sectors}$$

$$L_i = \text{mean relative sector time in late sectors}$$

Useful derived features:

$$\text{late_vs_early}_i = L_i - E_i$$

or:

$$\text{late_early_ratio}_i = \frac{L_i}{E_i}$$

Interpretation:

- High value: runner slowed down later relative to early race.
- Low value: runner finished stronger relative to early race.

## Positive Split / Negative Split

A simple pacing concept:

- Positive split: second half slower than first half.
- Negative split: second half faster than first half.

For long trail races, true negative splits may be rare because terrain changes, fatigue,
and course profile matter. So we should compute this after sector/course normalization,
not only using raw time.

## Pacing Slope

Fit a simple line to the runner's normalized sector pace over sector number:

$$r_{ij} = \alpha_i + \gamma_i \cdot j + e_{ij}$$

where:

- $j$ is sector number.
- $r_{ij}$ is normalized sector time.
- $\gamma_i$ is the pacing slope.

Interpretation:

- $\gamma_i > 0$: runner becomes relatively slower later.
- $\gamma_i < 0$: runner becomes relatively faster later.
- $\gamma_i \approx 0$: runner is relatively steady.

This gives one compact number for pacing trend.

## Pacing Variability

Measure how uneven a runner's pacing is:

$$V_i = \text{sd}(r_{i1}, r_{i2}, ..., r_{iK})$$

Interpretation:

- Low variability: steady pacing relative to comparable runners.
- High variability: uneven pacing, possibly surges, crashes, long stops, or terrain
  mismatch.

## Early Aggression

Define early aggression as being faster than expected in the early part of the race.

For example:

$$A_i = 1 - E_i$$

where $E_i$ is early relative sector time.

Interpretation:

- Larger $A_i$: runner started faster relative to their group.
- Smaller or negative $A_i$: runner started more conservatively.

We can then test whether high early aggression predicts worse later outcomes.

## Fade / Blow-Up Score

A "fade" score can compare late-race relative pace against early-race relative pace:

$$F_i = L_i - E_i$$

Interpretation:

- $F_i > 0$: runner faded later.
- $F_i < 0$: runner improved later.

This should be one of the central features.

# DNF Handling

DNF runners are very important, but they should probably not be mixed blindly with
finishers in final-time models.

There are at least three useful DNF analyses.

## Analysis A: Finishers Only

First analyze only finishers. This gives the cleanest final-time outcome:

$$T_i = \text{final race time}$$

This is the best first step because all runners have complete sector paths.

Limitation:

This ignores the worst outcomes, and pacing mistakes may be strongly connected to DNF.

## Analysis B: DNF Probability

Create a binary outcome:

$$D_i =
\begin{cases}
1, & \text{runner DNF} \\
0, & \text{runner finished}
\end{cases}
$$

Then model:

$$P(D_i = 1) = f(\text{ability}, \text{early pacing}, \text{early variability}, ...)
$$

This asks:

> Does starting too fast increase DNF probability among runners of similar ability?

Important detail:

Only use features available before the DNF point. For example, we can use first 25% or
first 50% of sectors, but we should not use late-race sectors that DNF runners never
reached.

## Analysis C: Survival / Dropout Timing

Treat each checkpoint as a point where the runner is still active or has dropped.

This becomes a time-to-event problem:

$$h_{ij} = P(\text{DNF at checkpoint } j \mid \text{still active before } j)
$$

This is more advanced, but it is conceptually a very good fit for ultramarathon data.
It asks:

> At which checkpoint does pacing stress start to predict dropout?

# Suggested First Workflow

## Step 1: Build A Clean Runner-Level Table

Create one row per `runner_id`.

Include:

- `race_id`
- `year`
- `runner_id`
- `runner_bib_number`
- `runner_overall_index`
- final status
- final time in minutes
- number of reached checkpoints
- DNF flag

This table is the base for group definitions and final outcomes.

## Step 2: Confirm Sector Structure

For each `race_id` and `year`, confirm:

- sector IDs are consecutive from `0` to max sector.
- every finisher has the expected number of sectors.
- DNF runners have fewer sectors or missing later sectors.
- sector times are numeric minutes.

Useful checks:

```{r}
# Pseudocode:
# sector_count_by_runner = df.groupby("runner_id")["sector_checkpoint_id"].nunique()
# sector_count_by_runner.describe()
```

## Step 3: Create Ability Groups

Start with 5 groups based on `runner_overall_index`.

Then check:

- number of runners per group
- final time distribution per group
- DNF rate per group

If groups are very imbalanced, use quantiles instead of fixed bins.

## Step 4: Create Normalized Pacing Features

For each runner-sector row:

- sector time share of total finish time
- sector time relative to ability-group median
- cumulative time relative to ability-group median

Then create runner-level summaries:

- early relative pace
- middle relative pace
- late relative pace
- late-vs-early fade score
- pacing slope
- pacing variability
- first-half vs second-half ratio

## Step 5: Analyze Finishers

For finishers only:

Outcome options:

- final time residual after accounting for `runner_overall_index`
- final time rank within ability group
- final time percentile within ability group

Core question:

> Among similar runners, do conservative starters, steady pacers, or late movers
> perform better?

Recommended first model:

$$\text{performance residual}_i =
\beta_0 +
\beta_1 \cdot \text{early aggression}_i +
\beta_2 \cdot \text{fade score}_i +
\beta_3 \cdot \text{pacing variability}_i +
\epsilon_i
$$

## Step 6: Analyze DNF Separately

For all starters:

Outcome:

$$D_i = 1 \text{ if DNF, else } 0
$$

Start with simple comparisons:

- DNF rate by ability group.
- DNF rate by early pacing aggressiveness.
- DNF rate by early sector rank relative to ability group.

Then build a simple model:

$$P(D_i = 1) =
f(\text{overall index}, \text{ability group}, \text{early aggression}, \text{early variability})
$$

Important:

Only use early-race features that are available for both finishers and DNF runners.

# Visual Ideas

Good visualizations for this project:

- Line chart of median normalized pacing profile by ability group.
- Line chart of normalized sector pace for top-performing vs under-performing runners.
- Heatmap of sector-relative performance by runner, sorted by final performance residual.
- Scatter plot: early aggression vs final performance residual.
- Scatter plot: fade score vs final performance residual.
- Boxplot: DNF rate or final residual across pacing strategy clusters.
- Checkpoint survival curve: percentage of runners still active by checkpoint and ability group.

# Pacing Strategy Clusters

After creating normalized pacing features, cluster runners into strategy types.

Possible clusters:

- fast starters / strong fade
- conservative starters / strong finish
- steady pacers
- volatile pacers
- late-race collapse

This should come after feature engineering, not before. Clustering raw sector times would
mostly rediscover ability differences. Clustering normalized pacing profiles is much
more meaningful.

# Possible Research Questions

Good concrete questions:

1. Among runners with similar `runner_overall_index`, which pacing profile produces
   the best final-time residual?
2. Is starting faster than ability-group median associated with worse late-race
   performance?
3. Is pacing variability associated with worse final outcome?
4. Do stronger runners tolerate aggressive starts better than weaker runners?
5. Are DNFs preceded by unusually aggressive early pacing?
6. At which sector does the difference between strong and weak outcomes become visible?
7. Are successful runners steady, or do they intentionally lose/gain time in specific
   course sections?

# Important Risks

## Course Profile Confounding

Raw sector time depends heavily on terrain. A slow sector may simply be uphill. This is
why sector-relative normalization is important.

## Ability Index Meaning

We need to confirm the direction and meaning of `runner_overall_index`. If the index is
not calibrated well, it should still be useful for grouping, but conclusions should be
phrased carefully.

## DNF Bias

Finishers-only analysis excludes the worst outcomes. DNF analysis should be separate
and should use only features available before dropout.

## Missing Data

Missing sector times, missing runner URLs, missing final times, or incomplete checkpoint
paths can all affect results. Each cleaning decision should be recorded.

## Same Runner Across Years

If we later include multiple years, the same real person may appear multiple times.
That is not necessarily a problem, but the unit should be clear:

- runner identity across all years
- runner-race-year identity

For pacing strategy, `runner_id` as runner/race/year is probably the right unit.

# Recommended First Analysis Version

The first clean version should be simple:

1. Finishers only.
2. One race and one year.
3. Ability groups based on `runner_overall_index` quantiles.
4. Sector times converted to minutes.
5. Sector IDs normalized to `0, 1, 2, ..., K`.
6. Pacing features:
   - early relative pace
   - middle relative pace
   - late relative pace
   - fade score
   - pacing variability
7. Outcome:
   - final time residual after predicting final time from `runner_overall_index`

Then compare:

$$\text{performance residual} \sim
\text{early relative pace} +
\text{fade score} +
\text{pacing variability}
$$

This gives a first answer to:

> After accounting for pre-race ability, what pacing behavior is associated with
> better race execution?

# Plain-English Summary

The cleanest way to think about the project is:

> Do not compare every runner directly to every other runner. First compare runners to
> runners with similar expected ability. Then ask whether their pacing shape explains
> who beat expectation, who underperformed, and who dropped out.

The central object should be a normalized pacing profile for each runner. Once each
runner has that profile, we can summarize it with simple features such as early
aggression, late fade, steadiness, and variability. Those features can then be compared
against final performance relative to expectation.

This keeps the analysis fair, interpretable, and close to the actual race story.
